\vskip-0.6in Trade Agreements in the Shadow of Lobbying\protect\footnote{Formerly circulated under the title ``Trade Agreements, Lobbying and Separation of Powers''}

### 2016-06-10 draft.tex
condense and move sec 2.4 text; closes #14; closes #15

\newpage
The fact that zero contributions are predicted in equilibrium is not particularly satisfying. The addition of political uncertainty provides the missing realism, smooths the results and delivers additional intuition. I assume all actors are uncertain about the weight the median legislator in a non-unitary legislature places on the profits of the lobbying industry. \textbf{Similar to probabilistic voting models that incorporate policy motivation such as Roemer (1997),} this represents the idea that the lobbying and voting ``game'' that goes on within legislatures is often complex enough that none of the actors know precisely what the outcome will be---that is, no one can predict exactly which legislator will be the median and therefore what weight will be placed on import-industry profits at the time of the vote.

@@

\subsection{Related Literature}
The foundations of this work rest on Grossman and Helpman's (1994) `Protection for Sale.' Two related papers `Trade Wars and Trade Talks' (Grossman and Helpman (1995b)) and `The Politics of Free-Trade Agreements' (Grossman and Helpman (1995a)) first employed the PFS approach in the study of trade agreements. The first considers ``Trade Wars'' and ``Trade Talks'' separately, whereas my approach allows the ``Trade Talks'' to be influenced by the desire to avoid a ``Trade War'' and thus views the ``Trade War'' as a crucial subgame. Their model of preferential trade agreements (Grossman and Helpman (1995a)), although treating the government as a unitary actor, is closer to this approach.

Another related literature considers the impact of exogenously-determined political uncertainty on the potential for trade cooperation. These studies (e.g. Feenstra and Lewis (1991); Milner and Rosendorff (2001); Bagwell and Staiger (2005); Beshkar (2010); Horn, Maggi and Staiger (2010); Beshkar and Bond (2012); Amador and Bagwell (2013)) derive various implications for the design of international trade agreements using a Baldwin (1987)-style government welfare function with exogenous shocks to the political-pressure parameter. As explained in the next section, the model presented here is intended to be a bridge between this literature and that discussed above that focuses on endogenously-determined political economy weights. To preview, this model is closest to that of Bagwell and Staiger (2005), with two main changes: in place of their unitary government signing a trade agreement and having different preferences ex-ante and ex-post, this model has two branches of government with differing preferences, where---in the case of the legislature---the political economy weights are determined endogenously.

@@

In the event that the trade agreement does not enter into or remain in force, there is a final stage of lobbying, resolution of the uncertainty surrounding this decision, and voting to set the trade-war tariffs.\footnote{Alternatively, one can model the lobbying for the ratification decision as determining the potential non-cooperative tariff; this changes behavior slightly depending on the relationship between the lobby and the legislature's preferences, but does not change the fundamental dynamics of the model. Yet another possibility is to replace the trade-war phase with the ex-ante status quo, which would most naturally be the non-cooperative equilibrium. This is precisely the outcome that obtains in the trade-war phase, so the results of the model are unaltered.} Once all political decisions are taken, producers and consumers make their decisions.

@@

I specify the welfare function of the home legislature as
\begin{equation}
  W_{\mathit{ML}} = \mathit{CS}_X(\tau) + \ga(e,\ve) \cdot \mathit{PS}_X(\tau) + \mathit{CS}_Y(\tau^*) + \mathit{PS}_Y(\tau^*) + \mathit{TR}(\tau)
  \label{eq:ml}
\end{equation}
where $\mathit{CS}$ is consumer surplus, $\mathit{PS}$ is producer surplus, $\ga(e,\ta)$ is the weight placed on producer surplus (profits) in the import-competing industry, $e$ is lobbying effort, $\ta$ is a random variable, and $\mathit{TR}$ is tariff revenue.\footnote{Labor income $l$ could also be included in both median legislator and executive welfare. I omit it because its inclusion alters none of the results and only serves to complicate the exposition.} I model the decisions of the legislature as being taken by a median legislator (more on this in Section~\ref{sec:choices}). Notice that the weight the median legislator places on the profits of the import-competing industry, $\ga(e,\ve)$, is affected by the level of lobbying effort and a random variable $\ta$.

@@

The home executive's welfare is specified as follows:
\[
  W_E = \mathit{CS}_X(\tau) + \ga_E \cdot \mathit{PS}_X(\tau) + \mathit{CS}_Y(\tau^*) + \mathit{PS}_Y(\tau^*) + \mathit{TR}(\tau)
\]
Note that this is identical to the welfare function for the legislature aside from the weight $\ga_E$ the executive places on the profits of the import industry, which is not a function of lobbying effort. Section~\ref{sec:choices} contains an in-depth discussion of these modeling choices.

@@

\subsection{Modeling Choices}
\label{sec:choices}

In modeling the objective of the legislature (see Equation~\ref{eq:ml}), I have introduced a variant of the general political support function introduced by Hillman (1982). While Hillman is credited with first proposing that policy makers choose tariffs in order to maximize their own political support, the PFS model contributed the explicit specification that lobbying contributions enter the political support function with constant marginal utility as $C + aW$.

Dixit, Grossman and Helpman (1997) point out that linearity in contributions prevents complete analysis of distributional questions; the PFS form also restricts the returns to lobbying activity to be constant.\footnote{See Section \ref{sec:gh} for further discussion of this point and an in-depth discussion of the relationship between the functional form presented herein and the PFS form. Note that the assumption of diminishing returns to lobbying effort has been present in the literature going back at least to Findlay and Wellisz (1982).} They show that the PFS equilibrium analysis can be extended to a more general setting where, among other features, returns to lobbying are not restricted to be constant.

Now, notice that, aside from the endogeneity of the weight the legislature places on the lobbying industry's profits, Equation~\ref{eq:ml} is precisely the \textit{deus ex machina} government objective function popularized by Baldwin (1987) that is commonly employed in the literature on the political economy of trade agreements. Instead of the weight on profits being constant as in the exogenous case or necessarily linear in contributions as in the PFS case, this form allows for both endogeneity and non-linearity.

Since trade policies are overwhelmingly determined within the context of trade agreements, it is useful to have a framework to bring together the endogenous political pressure of PFS-style modeling with the trade agreements approach. The formulation in Equation~\ref{eq:ml} is intended as a bridge between the two. See Buzard (2016) for an example of how letting the weight the median legislator places on the profits of the import-competing industry $\ga(e,\ve)$ be affected by the level of lobbying effort as well a random variable $\ta$ can integrate both endogenous political activity and exogenous shocks into the analysis of questions regarding optimal trade agreement design and the political welfare of governments.

Modeling the objective function so closely on the standard in the trade agreements literature allows for direct comparisons to the large extant body of work that studies exogenous shocks only, revealing cleanly the effects of the addition of endogenous lobbying---and if one wants---this different kind of political uncertainty.

While a structural model of the legislative process might be more satisfying, the reduced form can deliver the key insights without distracting from the central policy choice and lobbying dynamics that take place at the level of the executives and lobbies, to which we turn next.

The executives' objective function is taken to be identical to the legislatures' except that profits in the import-competing industry receive a fixed, exogenous weight. This construction permits me to integrate the influence of the import-competing industry on negotiations prior to the formation of the trade agreement while also reflecting the idea that the mechanisms through which political considerations influence the executives' preferences appear to be quite different from those at the legislative level---the electoral college and primary/caucus systems for instance.

This setup does not strictly require that the executives are not lobbied. The assumption that the weight the executives place on import-competing profits is constant only requires that the extent to which they favor this industry cannot be \textit{directly} influenced by lobbying effort. The assumption is innocuous as long as the executives' preferences are not directly altered in a significant way by lobbying over trade.

In the case of the post-war United States, where the Congress has consistently been significantly more protectionist than the President, this seems to reasonably reflect the political reality. For trade policy, where there are concentrated benefits but harm is diffuse, there are good reasons for this to be the case. Because the President has the largest constituency possible, delegating authority to the executive branch may simply be a mechanism for ``concentrating'' the benefits since consumers seem unable to overcome the free-riding problem. In fact, a strong argument can be made that power over trade policy has been delegated to the executive branch precisely \textit{because} it is less susceptible to the influence of special interests (Destler (2005)).

Therefore, in line with both the theoretical and empirical literature, I make one more assumption on $\ga(e,\ta)$:

That is, even for the least favorable outcome of the lobbying process, the legislature will be at least slightly more protectionist than the executive regardless of the lobby's effort choice. This assumption is not essential, but it matches well the empirical findings that politicians with larger constituencies are less sensitive to special interests, and it simplifies the analysis considerably, in particular ensuring that trade agreement tariffs are less than trade war tariffs (See Destler (2005) and footnote~\ref{fn:ga_l_e} above).

Although the political process here most closely matches that of the United States in the post-war era,\footnote{In particular, the binary decision by the legislature about whether to abide by or break the trade agreement is modeled on the ``Fast Track Authority'' that the U.S. Congress granted to the Executive branch almost continuously from 1974-1994 and then again as ``Trade Promotion Authority'' from 2002-2007.} the model is broadly applicable because the central dynamic that emerges---that trade agreements are formed in the \textit{shadow} of lobbying pressure instead of in direct response to it---appears to be present in a wide range of countries in which the formation and maintenance of trade policy requires \textit{approval} by a politically susceptible body either at home or in a trading partner.

We can see that the results apply beyond separation-of-powers presidential systems because the presence of veto points is what is essential for the dynamic posited here to occur. Although the number of formal veto points tends to be lower in, for instance, parliamentary systems, veto points are typically not absent altogether, at least in democracies (Henisz and Mansfield 2006). In reality, constraints on the executive are almost always present in democratic regimes, although they may operate in different ways, e.g. over questions of government dissolution and formation in parliamentary regimes. It is important to consider that constraints may be formal or informal, and they may come from legislative, regulatory, judicial, sub-national or even other executive bodies.

%I believe the one major fudge between my model and Destler's evidence is acceptable: that the executive is actually the one who breaks trade agreements, but they do it largely to prevent the legislature from doing something far worse.

>>>>

The fact that zero contributions are predicted in equilibrium is not particularly satisfying. The addition of political uncertainty provides the missing realism, smooths the results and delivers additional intuition. I assume all actors are uncertain about the weight the median legislator in a non-unitary legislature places on the profits of the lobbying industry. \textbf{Similar to probabilistic voting models that incorporate policy motivation such as Roemer (1997),} this represents the idea that the lobbying and voting ``game'' that goes on within legislatures is often complex enough that none of the actors know precisely what the outcome will be---that is, no one can predict exactly which legislator will be the median and therefore what weight will be placed on import-industry profits at the time of the vote.

@@

\textbf{Although the political process here most closely matches that of the United States in the post-war era,\footnote{In particular, the binary decision by the legislature about whether to abide by or break the trade agreement is modeled on the ``Fast Track Authority'' that the U.S. Congress granted to the Executive branch almost continuously from 1974-1994 and then again as ``Trade Promotion Authority'' from 2002-2007 and 2015 to present.} the model is broadly applicable because the central dynamic that emerges---that trade agreements are formed in the \textit{shadow} of lobbying pressure instead of in direct response to it---appears to be present in a wide range of countries in which the formation and maintenance of trade policy requires approval by a politically susceptible body either at home or in a trading partner.}

\textbf{The results apply beyond separation-of-powers presidential systems because the presence of veto points is what is essential for the dynamic posited here to occur. Although the number of formal veto points tends to be lower in, for instance, parliamentary systems, veto points are typically not absent altogether, at least in democracies (Henisz and Mansfield 2006). In reality, constraints on the executive are almost always present in democratic regimes, although they may operate in different ways.}

\subsection{Related Literature}
The foundations of this work rest on the `political support function' introduced by Hillman (1982) and Grossman and Helpman's (1994) `Protection for Sale.' Two related papers `Trade Wars and Trade Talks' (Grossman and Helpman (1995b)) and `The Politics of Free-Trade Agreements' (Grossman and Helpman (1995a)) first employed the PFS approach in the study of trade agreements. The first considers ``Trade Wars'' and ``Trade Talks'' separately, whereas my approach allows the ``Trade Talks'' to be influenced by the desire to avoid a ``Trade War'' and thus views the ``Trade War'' as a crucial subgame. Their model of preferential trade agreements (Grossman and Helpman (1995a)), although treating the government as a unitary actor, is closer to this approach.

Another related literature considers the impact of exogenously-determined political uncertainty on the potential for trade cooperation. These studies (e.g. Feenstra and Lewis (1991); Milner and Rosendorff (2001); Bagwell and Staiger (2005); Beshkar (2010); Horn, Maggi and Staiger (2010); Beshkar and Bond (2012); Amador and Bagwell (2013)) derive various implications for the design of international trade agreements using a Baldwin (1987)-style government welfare function with exogenous shocks to the political-pressure parameter.

\textbf{Since trade policies are overwhelmingly determined within the context of trade agreements, it is useful to have a framework to bring together the endogenous political pressure of PFS-style modeling with the trade agreements approach. The government objective function introduced in this paper is intended to be a bridge between the two. Buzard (2016) is an example of how this modeling approach can be used to integrate  endogenous political activity and exogenous shocks into the analysis of questions regarding optimal trade agreement design.}

\textbf{The objective function I introduce starts from the \textit{deus ex machina} government objective function popularized by Baldwin (1987). It is more flexible in that the weight the legislature places on profits is endogenous (PFS) as well as potentially non-linear (Findlay and Wellisz 1982); Dixit, Grossman and Helpman 1997) and stochastic.}

\textbf{Modeling the objective function so closely on the standard in the trade agreements literature allows for direct comparisons to the large extant body of work that studies exogenous shocks only, revealing cleanly the effects of the addition of endogenous lobbying---and if one wants---this different kind of political uncertainty.} To preview, this model is closest to that of Bagwell and Staiger (2005), with two main changes: in place of their unitary government signing a trade agreement and having different preferences ex-ante and ex-post, this model has two branches of government with differing preferences, where---in the case of the legislature---the political economy weights are determined endogenously.

@@

In the event that the trade agreement does not enter into or remain in force, there is a final stage of lobbying, resolution of the uncertainty surrounding this decision, and voting to set the trade-war tariffs.\footnote{Alternatively, one can model the lobbying for the ratification decision as determining the potential non-cooperative tariff; this changes behavior slightly but does not change the fundamental dynamics. Yet another possibility is to replace the trade-war phase with the ex-ante status quo, which would most naturally be the non-cooperative equilibrium. This is precisely the outcome that obtains in the trade-war phase, so the results of the model are unaltered.} Once all political decisions are taken, producers and consumers make their decisions.

@@

I specify the welfare function of the home legislature as
\begin{equation}
  W_{\mathit{ML}} = \mathit{CS}_X(\tau) + \ga(e,\ve) \cdot \mathit{PS}_X(\tau) + \mathit{CS}_Y(\tau^*) + \mathit{PS}_Y(\tau^*) + \mathit{TR}(\tau)
  \label{eq:ml}
\end{equation}
where $\mathit{CS}$ is consumer surplus, $\mathit{PS}$ is producer surplus, $\ga(e,\ta)$ is the weight placed on producer surplus (profits) in the import-competing industry, $e$ is lobbying effort, $\ta$ is a random variable, and $\mathit{TR}$ is tariff revenue.\footnote{Labor income $l$ could also be included in both median legislator and executive welfare. I omit it because its inclusion alters none of the results and only serves to complicate the exposition.} I model the decisions of the legislature as being taken by a median legislator. Notice that the weight the median legislator places on the profits of the import-competing industry, $\ga(e,\ve)$, is affected by the level of lobbying effort and a random variable $\ta$. I make the following assumptions on $\ga(e,\ta)$:

@@

The home executive's welfare is specified as follows:
\[
  W_E = \mathit{CS}_X(\tau) + \ga_E \cdot \mathit{PS}_X(\tau) + \mathit{CS}_Y(\tau^*) + \mathit{PS}_Y(\tau^*) + \mathit{TR}(\tau)
\]
Note that this is identical to the welfare function for the legislature aside from the weight $\ga_E$ the executive places on the profits of the import industry, which is not a function of lobbying effort. \textbf{This setup does not strictly require that the executives are not lobbied, only that the extent to which the executives favor this industry cannot be \textit{directly} influenced by lobbying effort. The assumption is innocuous as long as the executives' preferences are not directly altered in a significant way by lobbying over trade.}

\textbf{I make one more assumption on $\ga(e,\ta)$:}

\textbf{That is, even for the least favorable outcome of the lobbying process, the legislature will be at least slightly more protectionist than the executive regardless of the lobby's effort choice. This assumption is not essential, but it simplifies the analysis considerably, ensuring that trade agreement tariffs are less than trade war tariffs. It matches the empirical findings that politicians with larger constituencies are less sensitive to special interests and seems to represent particularly well the post-war U.S. where Congress has consistently been significantly more protectionist than the President (See Destler (2005) and footnote~\ref{fn:ga_l_e} above).}

@@

### 2016-06-10 letter-round2.tex
condense and move sec 2.4 text; closes #14; closes #15

\begin{itemize}
\renewcommand\labelitemi{-}
	\item \textit{Avoid changes of notation as you go along. This is totally unhelpful to the reader. Please pick one and stick to it throughout the paper.}
	\item \textit{Avoid digressions, lengthy discussions, and overreach: you stretch your model far too much. Stick to the main story. The paper still degenerates far too often into a series of discussions that bring little additional insight.}
	\item \textit{Section 2.4 either belongs to section 1.1 or should move as section 1.2. Section 2.4 is far too long.}
	\item \textit{Your recall stuff far too often. Not all readers suffer from memory losses.}
\end{itemize}

>>>>

\begin{itemize}
\renewcommand\labelitemi{-}
	\item \textit{Avoid changes of notation as you go along. This is totally unhelpful to the reader. Please pick one and stick to it throughout the paper.}
	\item \textit{Avoid digressions, lengthy discussions, and overreach: you stretch your model far too much. Stick to the main story. The paper still degenerates far too often into a series of discussions that bring little additional insight.}
	\item \textit{Section 2.4 either belongs to section 1.1 or should move as section 1.2. Section 2.4 is far too long.} \\
	The text of Section 2.4 has been shorted and redistributed. I inserted the few paragraphs that concerned assumptions into Section 2.2 (preferences) since they could not come before the model was introduced. The rest (in condensed form) integrated nicely into the introduction and literature review. I have bolded them where inserted.

	\item \textit{Your recall stuff far too often. Not all readers suffer from memory losses.}
\end{itemize}
